Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-12 paper-new/discussion.tex
revisions to discussion section

\eric{I moved this from the intro. We may need to revise it.}

@@

Current approaches to constructing languages that satisfy the
graduality property include the methods of Abstracting Gradual Typing
\cite{garcia-clark-tanter2016} and the formal tools of the Gradualizer
\cite{cimini-siek2016}.  These allow the language developer to start
with a statically typed language and derive a gradually typed language
that satisfies the graduality. The main downside to these
approaches lies in their inflexibility: since the process in entirely
mechanical, the language designer must adhere to the predefined
framework.  Many gradually typed languages do not fit into either
framework, e.g., Typed Racket \cite{tobin-hochstadt06,
  tobin-hochstadt08} and the semantics produced is not always the
desired one.
%
Furthermore, while these frameworks do prove graduality of the
resulting languages, they do not show the correctness of the
equational theory, which is equally important to sound gradual typing.

@@

\subsection{Synthetic Ordering}
\max{cut this subsection if we need space}
A key to managing the complexity of our concrete construction is in
using a \emph{synthetic} approach to step-indexing rather than working
analytically with presheaves. This has helped immensely in our ongoing
mechanization in cubical Agda as it sidesteps the need to formalize
these constructions internally. 
%
However, there are other aspects of the model, the bisimilarity and
the monotonicity, which are treated analytically and are similarly
tedious.
%
It may be possible to utilize further synthetic techniques to reduce
this burden as well, and have all type intrinsically carry a notion of
bisimilarity and ordering relation, and all constructions to
automatically preserve them.
%
A synthetic approach to ordering is common in (non-guarded) synthetic
domain theory and has also been used for synthetic reasoning for cost
models \cite{fiore_1997,GrodinNSH24}.

@@

\max{Expand on this, e.g. what some of the challenges might be and what would be reusable}
In the future, we plan to apply our approach to give a denotational
semantics for languages that feature higher-order state or
runtime-extensible dynamic typing
\cite{DBLP:journals/corr/abs-2210-02169} as well as richer type
disciplines such as gradual dependent types and effect systems.

>>>>

% \eric{I moved this from the intro. We may need to revise it.}

@@

The methods of Abstracting Gradual Typing \cite{garcia-clark-tanter2016} and the
formal tools of the Gradualizer \cite{cimini-siek2016} provide a way to derive
from a statically-typed language a gradually-typed language languages that
satisfies graduality by construction. The main downside to these approaches lies
in their inflexibility: since the process in entirely mechanical, the language
designer must adhere to the predefined framework.  Many gradually typed
languages do not fit into either framework, e.g., Typed Racket
\cite{tobin-hochstadt06, tobin-hochstadt08} and the semantics produced is not
always the desired one.
%
Furthermore, while these frameworks do prove graduality of the resulting
languages, they do not show the correctness of the equational theory, which is
equally important to sound gradual typing.

@@

% \subsection{Synthetic Ordering}
% \max{cut this subsection if we need space}
% A key to managing the complexity of our concrete construction is in
% using a \emph{synthetic} approach to step-indexing rather than working
% analytically with presheaves. This has helped immensely in our ongoing
% mechanization in cubical Agda as it sidesteps the need to formalize
% these constructions internally. 
% %
% However, there are other aspects of the model, the bisimilarity and
% the monotonicity, which are treated analytically and are similarly
% tedious.
% %
% It may be possible to utilize further synthetic techniques to reduce
% this burden as well, and have all type intrinsically carry a notion of
% bisimilarity and ordering relation, and all constructions to
% automatically preserve them.
% %
% A synthetic approach to ordering is common in (non-guarded) synthetic
% domain theory and has also been used for synthetic reasoning for cost
% models \cite{fiore_1997,GrodinNSH24}.

@@

% \max{Expand on this, e.g. what some of the challenges might be and what would be reusable}
% In the future, we plan to apply our approach to give a denotational
% semantics for languages that feature higher-order state or
% runtime-extensible dynamic typing
% \cite{DBLP:journals/corr/abs-2210-02169} as well as richer type
% disciplines such as gradual dependent types and effect systems.

% \max{Expand on this, e.g. what some of the challenges might be and what would be reusable}
In the future, we plan to apply our approach to give a denotational semantics
for languages that feature higher-order state or runtime-extensible dynamic
typing \cite{DBLP:journals/corr/abs-2210-02169} as well as richer type
disciplines such as gradual dependent types and effect systems. This will
involve adapting prior operational models based on step-indexing (e.g.,
\cite{new-giovannini-licata-2022}) to the denotational setting and using guarded
type theory to obtain solutions to guarded domain equations as we did in this
work for the denotation of the dynamic type. The generality of our techniques
should allow for many of the constructions to be reused across different
languages. For instance, the free error domain construction can be easily
extended with additional cases to model effects besides error and stepping.
Additionally, we aim to complete our Agda formalization and evolve it into a
reusable framework for mechanized denotational semantics of gradually-typed
languages.
